Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-07-08 analysis/paper/paper.qmd
revise the discussion section

# Discussion
<!--results and chronology -->
The faunal assemblage from the major residential area at Kiwulan suggests a continuous reliance on sika deer from CL6 to CL2, followed by a shift to bovine in CL1. Despite deer were the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence. The cultural layers were grouped into four temporal sequences in relation to foreign presences for the following interpretation. The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1). The term "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan. Although the small sample size across periods limits comprehensive interpretation, examining results in a continuous sequence still reveals distinct patterns in animal use.

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity. This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period. In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods. The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer. Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal]. A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods. Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods. Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period. Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer. Changes in cooking practices are also evident. The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997]. The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices. This could be further confirmed through organic residue analysis on pottery. Other meat preparation methods, such as drying, are also possible [e.g., @Spyrou2019]. Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963]. Moreover, some features containing long, thin (~4 cm in diameter) or Y-shaped racks may have been associated with drying facilities for meat or hides [@Chen2007; @Friesen2001]. The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.

A consistent pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs was identified across all periods. This might suggest selective butchering and a preference for meaty portions. Moreover, among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented due to marrow extraction, as these bones have high marrow yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Evidence for marrow extraction is supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation overall. Another explanation is that some hindquarters were used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs. If deer legs, particularly hindquarters, were traded as bone-meat packages, this could account for the lower frequencies of hindquarter bones compared to front quarters. While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat might be not only used for local consumption but also processed for trade. Thus, both factors could contribute to this pattern, rather than a single factor alone.

<!--deer hides-->
Deerskin trade had been prevalent in East Asia since the 14th century and was monopolized by Europeans in the 17th century when a large number of hides, especially from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981]. The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins, likely for trade with external traders. Moreover, the abundance of foreign goods at Kiwulan, including ornaments, porcelains, and metal items, further supports the frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007]. This suggests that northeastern Taiwan, like western Taiwan, was actively involved in and impacted by long-distance trade. Although lower limbs with skinning cuts did not show a substantial increase during the European period, the disproportionate rise in head parts might hint at the likelihood of brain tanning when preparing hides. Brain tanning is a traditional method for softening and preserving hides in many Indigenous cultures, including those of North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. In Taiwan, the present-day Bunun group in the central highlands has revived hide tanning based on oral history, primarily using peanuts or pig lard as tanning materials, with occasional use of animal brains [@Li1999, pp. 48; @Xu1994]. Despite the specific tanning materials used by the Kavalan people remain unknown due to the loss of traditions, brain tanning is a plausible method given the availability of deer heads from hunting.

<!-- population dynamics and arguments-->
The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods. Third, population growth would have likely led to a more proportional increase in both large and small animals. Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion. This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and similar return rates to deer [@Broughton2011prey].

The role of suids as a secondary resource after deer has been observed in the Indigenous cultures in southwestern Taiwan during the Neolithic period and Iron Age [@Chuang2021], suggesting that suids may have served as a supplementary resource while deer remained the prioritized prey. At Kiwulan, it is possible that portions of deer were reserved for trade, limiting local consumption and leading to an increased reliance on suids. Domestic pigs were identified at Kiwulan in both the European and pre-European periods based on their much smaller third molar size (length < 30 mm) and morphology, in comparison to archaeological data from Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021]. However, a clear distinction between wild boars and domestic pigs in archaeological remains is challenging due to overlapping size ranges, regional variations, and the potential presence of hybrids [@Rowley2012; @Chuang2021]. The third molar lengths of suids found at Kiwulan (28–36 mm) may also reflect such hybridization patterns. While current evidence does not clearly indicate the emergence of intensified suid management, an increase in suid utilization, either through domestication or hunting of wild individuals, is evident during the European period.

After the Dutch left, the deerskin trade continued but on a much smaller scale, with around 10,000 pieces exported annually in the 18th century, which is far fewer than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011]. The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period. It should be noted that muntjac were consistently part of local subsistence, though in small numbers, until the early Chinese period. Their skins were used by many Indigenous groups such as the Tsou and Bunun for making clothing, similar to the use of deerskins [@Li1999], which may have also been practiced at Kiwulan. Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.

@@

# Conclusion
By situating zooarchaeological evidence within the context of foreign presence, the Europeans in the 17th century and Chinese immigrants in the 19th century, this study reveals Indigenous participation in regional trade networks and their adaptation strategies. The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period. Deer as a primary animal resource at Kiwulan suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products. Additionally, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer. These findings highlight how Kiwulan residents navigated foreign influence and environmental changes while maintaining cultural traditions. Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with relatively weak colonial control were profoundly impacted by colonial activities than previously thought. Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.

>>>>

# Discussion
<!--results and chronology -->
The faunal assemblage from the major residential area at Kiwulan suggests a continuous reliance on sika deer from CL6 to CL2, followed by a shift to bovine in CL1. Despite deer being the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence. The cultural layers were grouped into four temporal sequences in relation to foreign presences for the following interpretation. The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1). The term "European period" here does not refer to direct colonial control over Kiwulan but the broad impact of colonial activities in northern Taiwan, including long-distance trade, tribute systems, and occasional attacks on Indigenous communities in Yilan. Although the small sample size across periods may limit the scope of interpretation, the consistent analysis of faunal remains from each period still provides valuable insights into changes in animal use.

<!--summary of overall observations across time periods -->
Deer is the predominant animal, mostly from the European period, followed by the late pre-European period. Interestingly, deer bones from the European period show the same ubiquity as in the late pre-European period but appear in twice the quantity. This suggests a regular deer consumption at the site, but an intensified use during the European period. In addition to deer, suids were also used more frequently during the European period compared to other phases. The dramatic drop in deer bones after the European period may be associated with declines in both the wild deer population and the Kiwulan population. A similar pattern of shrunken sika deer populations has been recorded in western Taiwan since the 18th century due to habitat loss and overhunting [@Chen2000animal]. A slight increase in muntjac remains in the post-European period suggests they may have served as a substitute, replacing suids as a secondary resource from earlier periods. Nevertheless, deer remained a primary source for local consumption at Kiwulan during the post-European and early Chinese periods.

<!--skeletal part frequencies and butchery patterns of deer-->
Although deer remained a major resource prior to the late Chinese period, skeletal parts of deer indicate changes in butchery practices during the European period. More complete carcasses and a substantial increase in fragmented head parts (cranium and mandible) in the European period suggest that more intact deer were transported to and processed at the site, with greater use of soft tissues such as brains and tongues. These findings are consistent with the mentions about deer tongue consumption and Indigenous group hunting in *Record of Formosa (Dongfan ji)* in 1603 by Chinese travelers [@Chen1603]. Group hunting may have become more frequent during the European period, enabling the transport of whole deer. Shifts in cooking practices are also evident. Burned bones were common in the pre-European period but decreased largely in the European period, suggesting a change from roasting to other cooking methods, such as boiling, an ideal practice for maximizing meat use and extracting grease [@Lupo1997]. Many chopped bones at Kiwulan can fit well within the dimensions of local ceramics, implying the likelihood of boiling. This could be further confirmed through organic residue analysis on pottery in the future. Other meat preparation methods, such as drying, should also be considered [e.g., @Spyrou2019]. Although skeletal elements, such as vertebrae and ribs [@Friesen2001; @Binford1978], typically associated with drying are present in low frequencies at the site, the consumption and exchange of dried meat by the Kavalan people were documented in Chinese historical records [@Chen1963]. Moreover, wooden racks in long strips (~ 4 cm in diameter) and Y-shaped forms found in piles at Kiwulan might be associated with facilities for drying meat or hides [@Chen2007; @Friesen2001]. Overall, the changes in butchering and cooking practices suggest an increased need for deer, with methods like boiling or drying being used to optimize animal use and extend meat preservation.

For deer skeletal parts, the pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs suggests selective butchering and a preference for meaty portions. It should be noted that among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented due to marrow extraction, as these bones have high marrow yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Evidence for marrow extraction is supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation overall. Another explanation is that some hindquarters could be used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs. If deer legs, particularly hindquarters, were traded as bone-meat packages, this can account for the lower frequencies of hindquarter bones compared to front quarters. While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat may have been not only consumed locally but also processed for trade.

<!--deer hides-->
According to documentary sources, deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981]. Archaeological evidence of hide preparation can be supported by cut marks found on the lower limbs of deer at Kiwulan. Moreover, foreign goods found at Kiwulan, such as glass beads, porcelains, and metal objects, might be the result of trade with outsiders [@Wang2020ornament; @Chen2007]. These findings indicate that northeastern Taiwan was also actively engaged in the long-distance trade networks stimulated by European involvement. Although the frequency of lower limbs with skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning in addition to serving as a nutritional resource. Brain tanning is a method for softening and preserving hides used by many Indigenous cultures across North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. In Taiwan, the use of peanuts or pig lard as tanning materials has been culturally revived by the present-day Bunun people in the central highlands based on oral history, with occasional use of animal brains [@Li1999, pp. 48; @Xu1994]. Although the tanning methods once practiced by the Kavalan people remain unclear, brain tanning is a plausible way, given the availability of deer heads from hunting and the large number of cranial fragments found at Kiwulan during the European period. However, further archaeological evidence is still needed to fully support this possibility.

<!-- population dynamics and arguments-->
Population growth inferred from the high frequencies of pottery during the European period may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009]. Nevertheless, the impact of foreign presence is evident in deer butchery patterns, relative proportions of animals, and animal ubiquity. First, lower limbs with skinning cuts appear in the late pre-European phase and increase during the European period, which aligns with the development of the deerskin trade in East Asia in the 17th century. Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected. In fact, fractured long bones are 10–30% less frequent during the European and the late pre-European periods compared to other periods. Third, population growth would normally lead to a proportional increase in both large and small animal remains. However, despite the high taxonomic diversity during the European period, only suid remains showed a substantial rise in ubiquity, quantity, and proportion. This indicates that the trade of deer products may have led Indigenous people to rely more heavily on suids as an alternative resource due to their comparable body size and similar return rates to deer [@Broughton2011prey].

Suids may have been treated as a secondary resource, while deer remained the prioritized. This pattern is not only observed at Kiwulan, but also reported at some Neolithic and Iron Age sites in southwestern Taiwan [@Chuang2021]. At Kiwulan, it is possible that particular deer parts were reserved for trade, which therefore limited local consumption and led to an increased reliance on suids during the European period. Among the suids, a few domestic pigs were identified in both the European and pre-European periods based on their small third molars (length < 30 mm) and dental morphology, when compared with reference data from other Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021]. This suggests that small-scale pig domestication was practiced at Kiwulan. However, because domestic pigs and wild boars are difficult to distinguish due to similar body sizes, regional morphological variation, and the potential for hybridization, the number of domestic pigs may be underestimated [@Rowley2012; @Chuang2021]. The third molar lengths of suids from Kiwulan mostly range from 28 mm to 36 mm, which may reflect such hybridization. While the current evidence does not clearly indicate intensified suid management, the increased utilization of suids, whether through domestication or the hunting of wild individuals, is evident during the European period.

The deerskin trade continued after the Dutch left, but it was much smaller. In the 18th century, about 10,000 pieces were exported annually, far less than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011]. The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period. It should be noted that muntjac were consistently part of local subsistence, though in small numbers, until the early Chinese period. Their skins were used by many Indigenous groups such as the Tsou and Bunun for making clothing, similar to the use of deerskins [@Li1999], which may have also been practiced at Kiwulan. Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.

Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1). One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan, which occurred around the same time as a large influx of Han Chinese immigrants in the 19th century. The occupation of land by Han settlers gradually led to Kavalan relocation to eastern Taiwan, leaving only a small Indigenous population in Yilan [@Zhan1995]. Moreover, bovines are believed to have been introduced by Han populations because they are absent from prehistoric contexts in Taiwan [@Chen2000animal]. Therefore, the predominance of bovines in CL1 may indicate both a decrease in the Indigenous population and a shift from hunting to livestock use introduced or intensified under Han influence. Similar cultural changes can also be seen in burial practices at Kiwulan during the Chinese period, when Indigenous people began using wooden containers for burials that resembled Chinese coffins [@Chen2007; @Chiu2004].

@@

# Conclusion
The zooarchaeological evidence at Kiwulan reveals Indigenous participation in regional trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century. The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period. Deer consumption at Kiwulan throughout most of time suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products. On the other hand, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer. These findings highlight the adaptation of Kiwulan residents to foreign impacts and environmental changes while maintaining cultural traditions. Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with weak colonial control were profoundly impacted by colonial activities than previously thought. Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.
